The Response Time of Toronto's Paramedic Service: How effective is the city's emergency medical care?

### 2023-02-03 data-simulation.qmd
General updates and cleaning
* Improved comments in several files
* Fixed formatting issue caused by HTML comments appearing above Quarto file metadata
* Added a new file to install dependencies
* Added new citations to the paper
* Updated readme.md

```{r}
simulated_emergent_care_data <-
  tibble(
    # Define the type of pre-hospital care: while the raw data has multiple values for this, we only use Emergency Requests for Service
    "Pre-Hospital Emergency Care Type" = sample(
      x = "Emergency Request for Service",
      size = 8,
      replace = TRUE
    ),

@@

    # Vary the data between 200,000 and 300,000
    "Actual" = sample(
      x = c(200000:350000),
      size = 8,
      replace = TRUE
    )
)

@@

```{r}
simulated_incident_data <-
  tibble(
    # Vary the date between January 2, 2021 and December 31, 2021
    "Dispatch_Time" = sample(
      seq(as.Date('2021/01/01'), as.Date('2021/12/31'), by="day"), 
      size = 1000,
      replace = TRUE
    ),

    # Randomly pick an incident type1000 times
    "Incident_Type" = sample(
      x = c(
        "Medical",
        "Motor Vehicle Accident",
        "Emergency Transfer",
        "Fire",
        "Airport Standby",
        "-"
      ),
      size = 1000,
      replace = TRUE
    ),

>>>>

```{r}
simulated_emergent_care_data <-
  tibble(
    # Define the type of pre-hospital care (we only use this one value)
    "Pre-Hospital Emergency Care Type" = sample(
      x = "Emergency Request for Service",
      size = 8,
      replace = TRUE
    ),

@@

    # Randomly select numbers between 200,000 and 300,000
    "Actual" = sample(
      x = c(200000:350000),
      size = 8,
      replace = TRUE
    )
)

@@

```{r}
simulated_incident_data <-
  tibble(
    # Randomly pick dates between January 2, 2021 and December 31, 2021
    "Dispatch_Time" = sample(
      seq(as.Date('2021/01/01'), as.Date('2021/12/31'), by="day"), 
      size = 1000,
      replace = TRUE
    ),

    # Randomly pick an incident type 1000 times
    "Incident_Type" = sample(
      x = c(
        "Medical",
        "Motor Vehicle Accident",
        "Emergency Transfer",
        "Fire",
        "Airport Standby",
        "-"
      ),
      size = 1000,
      replace = TRUE
    ),

### 2023-02-03 inputs/data/data_import.qmd
General updates and cleaning
* Improved comments in several files
* Fixed formatting issue caused by HTML comments appearing above Quarto file metadata
* Added a new file to install dependencies
* Added new citations to the paper
* Updated readme.md

<!--
#### Preamble ####
# Purpose: Import data from the Toronto Paramedic Service via Open Data Toronto
# Author: Sebastian Rodriguez
# Email: me@srod.ca
# Date: 3 February 2023
# GitHub: https://github.com/seb646/toronto-paramedic-responses
-->
---
title: "Import data from the Toronto Paramedic Service via Open Data Toronto"
format: html
author: "Sebastian Rodriguez"
---
```{r}
#### Workspace set-up ####
library(opendatatoronto)
library(tidyverse)
library(dplyr)
```

>>>>

---
title: "Import data from the Toronto Paramedic Service via Open Data Toronto"
format: html
author: "Sebastian Rodriguez"
---
```{r}
#### Preamble ####
# Purpose: Import data from the Toronto Paramedic Service via Open Data Toronto
# Author: Sebastian Rodriguez
# Email: me@srod.ca
# Date: 3 February 2023
# GitHub: https://github.com/seb646/toronto-paramedic-responses
```

```{r}
#### Workspace set-up ####
library(opendatatoronto)
library(tidyverse)
library(dplyr)
```

### 2023-02-03 outputs/paper/analysis.qmd
General updates and cleaning
* Improved comments in several files
* Fixed formatting issue caused by HTML comments appearing above Quarto file metadata
* Added a new file to install dependencies
* Added new citations to the paper
* Updated readme.md

---
title: "The Response Time of Toronto's Paramedic Service: How effective is the city's emergency medical care?"
format: pdf
author: "Sebastian Rodriguez^[University of Toronto, Faculty of Information]"
abstract: "The Toronto Paramedic Services provides all pre-hospital care to the city’s inhabitants. Recent criticism of the service called into question its performance, highlighting staff shortages and institutional challenges that the city's paramedics face. This paper investigates the Toronto Paramedic Services’ efficacy by comparing its emergency response times to the standards set by provincial and municipal governments. It reveals that the Toronto's paramedics meet and often exceed response time standards, and provided stroke and STEMI heart attack patients with a chance of survival greater than the provincial average." 
date: today
date-format: long
editor: visual
toc: TRUE
number-sections: TRUE
bibliography: references.bib
thanks: "Code and data supporting this analysis is available at: https://github.com/seb646/toronto-paramedic-responses"
---

@@

```{r}
#| messages: false
#| echo: false
#| warning: false
# Define required libraries 
library(lubridate)
library(tidyverse)
library(dplyr)
library(knitr)
library(janitor)
library(scales)
library(RColorBrewer)
library(ggplot2)
library(kableExtra)
library(here)
library(readr)

# Fetch the .csv file created from Open Data Toronto's "Land Ambulance Response Time Standard" package: https://open.toronto.ca/dataset/land-ambulance-response-time-standard
raw_response_data <- readr::read_csv(here::here("inputs/data/raw_response_data.csv"), show_col_types = FALSE) |>
  clean_names()

# Fetch Open Data Toronto's "Pre-Hospital Emergency Care Performance Metrics" package:  https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics
raw_emergent_care_data <- readr::read_csv(here::here("inputs/data/raw_emergent_care_data.csv"), show_col_types = FALSE)

@@

Before investigating the response times and efficacy of the Toronto Paramedic Services, it is important to understand the service's breadth of responsibility. With a population of nearly 3 million people, Toronto requires a medical response to roughly 300,000 emergency calls every year. As the city's population grows, so too does the need for medical assistance. There was a steady increase of approximately 10,000 emergency medical requests per year from 2014 to 2019, as outlined by the graph below (@fig-1). This trend ended briefly in 2020 when the COVID-19 pandemic hit Toronto, placing an incredible strain on the city's healthcare system. Due to a concern about COVID-19 infection, many people delayed or avoided medical care throughout the pandemic [@cite_avoiding_hospitals].

```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Number of Emergency Medical Responses in Toronto (by year since 2014)"
#| label: fig-1
raw_emergent_care_data |>
  # Define plot 
  ggplot(aes(
    x = Year, 
    y = Actual
  )) +

@@

\newpage

To further illustrate the responsibility of Toronto's paramedics, the table below (@fig-2) outlines the number of incidents, and ambulance dispatches the service fielded in 2021. The paramedic service responded to over 310,000 calls, ranging from routine medical emergencies to motor vehicle accidents and fires. These calls required a total of 315,788 ambulances, with an average of 865 ambulances dispatched per day. Most calls and incidents need between 1 and three ambulances. However, some severe incidents placed an even greater strain on the service's resources:

@@

# Render table
clean_incident_data |>
  knitr::kable(align = "l", booktabs = TRUE, escape = FALSE, linesep = "") |>
  add_header_above(
    c(" ", "Year Totals" = 2, "Dispatch Averages" = 2), 
    align = "c",
    bold=TRUE
  ) |>
  kable_styling(font_size = 10, latex_options = "hold_position", full_width = TRUE) |>
  row_spec(0,bold=TRUE) |>
  column_spec(1, width = "13em") 
```

\newpage

@@

In Ontario, the *Ambulance Act* defines the rules governing paramedics' response times. The act defines response times as the difference between the time an emergency service dispatcher receives a call and the on-scene arrival of either an ambulance or a person equipped with a defibrillator [@cite_ambulance_act]. The *Ambulance Act* mandates varying response times based on the severity of a patient's injury. To define these response times and categorize patients based on their injuries, paramedics use the Canadian Triage Acuity Scale (CTAS). CTAS has 5 levels of severity, with level 1 representing life-threatening injuries and level 5 representing non-urgent injuries.

Specific response times are defined under Regulation 257/00, Part VIII of the *Ambulance Act*. This regulation sets the response times for patients suffering from Sudden Cardiac Arrest (SCA) or a CTAS Level 1 injury as 6 minutes and 8 minutes, respectively (source; source). CTAS Levels 2 through 5 do not have mandated response times in this regulation, leaving municipalities to define their response times based on the needs of their inhabitants and the abilities of their paramedic service. The table below (@fig-3) outlines the City of Toronto's response times for each CTAS level.

@@

It is important to note that not all calls are mandated to receive a response time within these parameters. Individual municipalities are responsible for defining a "performance plan," which determines the percentage of calls that must be covered by their response times. In Toronto, the enacted performance plan requires 75% of all calls to receive a response within the required time frame. However, prior to 2016, only 60% of SCA calls required a response within the mandated time. \newpage

@@

The Toronto Paramedic Services has an excellent track record of achieving, and often exceeding, the percentage of calls receiving mandated response times. The graphs below (@fig-4) chart the percentage of incidents served by the mandated response times versus the city's set performance plan. To place these graphs in perspective, the City of Toronto's paramedic service averaged a response time of 6 minutes and 58 seconds for "emergency calls" in 2021 [@cite_ambulance_program].

@@

There was only one year when Toronto's paramedics could not meet all of the requirements set by the city's performance plan. In 2013, only 70% of patients identified as CTAS Level 1 were served by the mandated response time, as opposed to the 75% coverage promised by the performance plan. The data provided to the public does not indicate specific response times for each case, so it is impossible to determine how much longer response times were compared to the mandates---in some cases, it may be a difference of a few seconds or more than a few minutes. \newpage

@@

## References
<!-- Span required to show references as a numbered section -->

>>>>

---
title: "The Response Time of Toronto's Paramedic Service: How effective is the city's emergency medical care?"
format: pdf
author: "Sebastian Rodriguez^[University of Toronto, Faculty of Information]"
abstract: "The Toronto Paramedic Services provides all pre-hospital care to the city’s inhabitants. Recent criticism of the service called into question its performance, highlighting staff shortages and institutional challenges that the city's paramedics face. This paper investigates the Toronto Paramedic Services’ efficacy by comparing its emergency response times to the standards set by provincial and municipal governments. It reveals that Toronto's paramedics meet and often exceed response time standards, and provide stroke and STEMI heart attack patients with a chance of survival greater than the provincial average." 
date: today
date-format: long
editor: visual
toc: TRUE
number-sections: TRUE
bibliography: references.bib
thanks: "Code and data supporting this analysis is available at: https://github.com/seb646/toronto-paramedic-responses"
---
```{r}
#| messages: false
#| echo: false
#| warning: false
#### Preamble ####
# Title: The Response Time of Toronto's Paramedic Service: How effective is the city's emergency medical care?
# Author: Sebastian Rodriguez
# Email: me@srod.ca
# Date: 3 February 2023
# GitHub: https://github.com/seb646/toronto-paramedic-responses
```

```{r}
#| messages: false
#| echo: false
#| warning: false
#### Workspace set-up ####
# Define required libraries 
library(lubridate)
library(tidyverse)
library(dplyr)
library(knitr)
library(janitor)
library(scales)
library(RColorBrewer)
library(ggplot2)
library(kableExtra)
library(here)
library(readr)
```

```{r}
#| messages: false
#| echo: false
#| warning: false
#### Fetch and clean relevant data sets ####
# Fetch the .csv file created from Open Data Toronto's "Land Ambulance Response Time Standard" package: https://open.toronto.ca/dataset/land-ambulance-response-time-standard
raw_response_data <- readr::read_csv(here::here("inputs/data/raw_response_data.csv"), show_col_types = FALSE) |>
  clean_names()

# Fetch Open Data Toronto's "Pre-Hospital Emergency Care Performance Metrics" package:  https://open.toronto.ca/dataset/pre-hospital-emergency-care-performance-metrics
raw_emergent_care_data <- readr::read_csv(here::here("inputs/data/raw_emergent_care_data.csv"), show_col_types = FALSE) |>
  clean_names()

@@

@@

Before investigating the response times and efficacy of the Toronto Paramedic Services, it is important to understand the service's breadth of responsibility. With a population of nearly 3 million people, Toronto requires a medical response to roughly 300,000 emergency calls every year. As the city's population grows, so too does the need for medical assistance. There was a steady increase of approximately 10,000 emergency medical requests per year from 2014 to 2019, as outlined by @fig-1 [@cite_emergency_care_data]. This trend ended briefly in 2020 when the COVID-19 pandemic hit Toronto, placing an incredible strain on the city's healthcare system. Due to a concern about COVID-19 infection, many people delayed or avoided medical care throughout the pandemic [@cite_avoiding_hospitals].

```{r}
#| messages: false
#| echo: false
#| warning: false
#| fig-cap: "Number of Emergency Medical Responses in Toronto (by year since 2014)"
#| label: fig-1
raw_emergent_care_data |>
  # Define plot 
  ggplot(aes(
    x = year, 
    y = actual
  )) +

@@

```
\newpage

To further illustrate the responsibility of Toronto's paramedics, @fig-2 outlines the number of incidents, and ambulance dispatches the service fielded in 2021 [@cite_paramedic_incident_data]. The paramedic service responded to over 310,000 calls, ranging from routine medical emergencies to motor vehicle accidents and fires. These calls required a total of 315,788 ambulances, with an average of 865 ambulances dispatched per day. Most calls and incidents need between 1 and three ambulances [@cite_paramedic_incident_data]. However, some severe incidents placed an even greater strain on the service's resources:

@@

# Render table
clean_incident_data |>
  knitr::kable(align = "l", booktabs = TRUE, escape = FALSE, linesep = "") |>
  add_header_above(
    c(" ", "Year Totals" = 2, "Dispatch Averages" = 2), 
    align = "c",
    bold=TRUE
  ) |>
  kable_styling(font_size = 10, latex_options = "hold_position", full_width = TRUE) |>
  row_spec(0,bold=TRUE) |>
  column_spec(1, width = "13em") 
```
\newpage

@@

In Ontario, the *Ambulance Act* defines the rules governing paramedics' response times. The act defines response times as the difference between the time an emergency service dispatcher receives a call and the on-scene arrival of either an ambulance or a person equipped with a defibrillator [@cite_ambulance_act]. The *Ambulance Act* mandates varying response times based on the severity of a patient's injury. To define these response times and categorize patients based on their injuries, paramedics use the Canadian Triage & Acuity Scale (CTAS). CTAS has 5 levels of severity, with level 1 representing life-threatening injuries and level 5 representing non-urgent injuries [@cite_ctas].

Specific response times are defined under Regulation 257/00, Part VIII of the *Ambulance Act*. This regulation sets the response times for patients suffering from Sudden Cardiac Arrest (SCA) or a CTAS Level 1 injury as 6 minutes and 8 minutes, respectively  [@cite_ambulance_act]. CTAS Levels 2 through 5 do not have mandated response times in this regulation, leaving municipalities to define their response times based on the needs of their inhabitants and the abilities of their paramedic service. @fig-3 outlines the City of Toronto's response times for each CTAS level [@cite_ambulance_response_data].

@@

It is important to note that not all calls are mandated to receive a response time within these parameters. Individual municipalities are responsible for defining a "performance plan," which determines the percentage of calls that must be covered by their response times [@cite_ambulance_act]. In Toronto, the enacted performance plan requires 75% of all calls to receive a response within the required time frame. However, prior to 2016, only 60% of SCA calls required a response within the mandated time [@cite_ambulance_response_data]. \newpage

@@

The Toronto Paramedic Services has an excellent track record of achieving, and often exceeding, the percentage of calls receiving mandated response times. The graphs below (@fig-4) chart the percentage of incidents served by the mandated response times versus the city's set performance plan [@cite_ambulance_response_data]. To place these graphs in perspective, the City of Toronto's paramedic service averaged a response time of 6 minutes and 58 seconds for "emergency calls" in 2021 [@cite_ambulance_program].

@@

There was only one year when Toronto's paramedics could not meet all of the requirements set by the city's performance plan. In 2013, only 70% of patients identified as CTAS Level 1 were served by the mandated response time, as opposed to the 75% coverage promised by the performance plan [@cite_ambulance_response_data]. The data provided to the public does not indicate specific response times for each case, so it is impossible to determine how much longer response times were compared to the mandates---in some cases, it may be a difference of a few seconds or more than a few minutes. \newpage

@@

## References
<!-- Comment required to show references as a numbered section -->
